%!TEX root = ../Main.tex

% \pagebreak
\section{Pipit core language}
\label{s:core}

\input{figures/core-structure.tex}

We now introduce the core Pipit language.
Note that this form differs slightly from the surface syntax presented earlier in \autoref{s:motivation}, which used the syntax of the metalanguage \fstar{}, as well as including proofs in \fstar{} itself.


\autoref{f:core:structure} shows the high-level architecture of Pipit.
On the left-hand-side, the surface syntax embedded in \fstar{} is shown; this includes some Pipit-specific syntactic sugar.
The translation from the surface syntax to the core language is trusted.
There are two targets from the core language: abstract transition systems for verification, and executable transition systems for extraction to C.
The translation to abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
The translation to abstract systems also generates proof obligations, which are verified to correspond to the proof obligations on the original program.
% One property that still remains to be proven is that the proof obligations on the abstract transition system entail the original proof obligations (\autoref{s:transition}); this pending proof is denoted as negated entailment in the figure ($\not\vdash$).
The translation to executable transition systems is proven to be semantics-preserving, as is the subsequent translation to \lowstar{}.
The translation from \lowstar{} to C is external to this paper and forms part of our trusted computing base.


\autoref{f:core-grammar} defines the grammar of Pipit.
The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and primitive applications ($p(\ov{e})$).
Most of the expression forms were introduced informally in \autoref{s:motivation} and correspond to the clock-free expressions of Lustre~\cite{caspi1995functional}.

\input{figures/core-grammar.tex}

The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
% Streams can also be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the values from stream $e'$ for subsequent steps.

Recursive streams are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
% To ensure that streams are productive,
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined and counts from zero up, but the stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
% We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \cite{caspi1995functional} which is not syntax-directed.
% Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax by combining the bindings together into a record or tuple.

Checked properties and contracts are annotated with their property status $\PStatus$, which can either be valid ($\PSValid$) or unknown ($\PSUnknown$).
For checked properies $\xcheckP{\PStatus}{e}$, the property status denotes whether the property has been proved to be valid.

Contracts $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ allow modular reasoning by replacing the implementation with an abstract specification.
Contracts involve two verification conditions.
Firstly, when a contract is \emph{defined}, the definer must prove that the body satisfies the contract: roughly, if $\erely$ is always true, then $\eguar[x := \ebody]$ is always true.
Secondly, when a contract is \emph{instantiated}, the caller must prove that the environment satisfies the precondition: that is, $\erely$ is always true.
Conceptually, then, a contract could have two property statuses: one for the definition and one for the instantiation.
However, in practice, it is not useful to defer the proof of a contract definition --- one could achieve a similar effect by replacing the contract with its implementation.
For this reason, we only annotate contracts with one property status, which denotes whether the instantiation has been proved to satisfy the precondition.

For example, the core expression $\sumrec$ computes the sum of values from a stream of integers \emph{ints} by defining a recursive stream $\sumvar$, which is delayed and given an initial value of zero.
If we were to use this sum in a context that required a strictly positive integer, we could give it a contract that states that if the input stream is always positive, then the resulting sum is also positive:
$$
\sumcontract
$$
To be considered a valid program, we must prove that the contract definition itself holds, as with our earlier contract (\autoref{s:motivation:contract}).
The unknown property status here allows us to defer the caller's proof that the input stream is always positive until the contract is used.


The remaining grammatical constructs of \autoref{f:core-grammar} describe streams, value environments, types and type environments.
Streams $V$ are represented as a sequence of values; streaming history environments $\Sigma$ are streams of heaps.
Types $\tau$ and type environments $\Gamma$ are standard.
For the presentation of the formal grammar here, we consider only a fixed set of values and primitives; in practice, the implementation is parameterised by a primitive table which we extend with immutable array operations for the TTCAN driver logic in \autoref{s:evaluation}.

\input{figures/core-typing.tex}

We define the typing judgments for Pipit in \autoref{f:core-typing}.
Most of the typing rules are standard for an unclocked Lustre.
The typing judgment $\typing{\Gamma}{e}{\tau}$ denotes that, in an environment of streams $\Gamma$, expression $e$ denotes a stream of type $\tau$.
This core typing judgment differs from the surface syntax used in \autoref{s:motivation}, which used an explicit stream type; for the core language, we instead assume that everything is a stream.

We use an auxiliary function $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$; for primitives $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ denotes that $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
Primitives are pure, non-streaming functions.

Rules \textsc{TValue}, \textsc{TVar}, \textsc{TPrim} and \textsc{TLet} are standard.

Rule \textsc{TFby} states that expression $\xfby{v}{e}$ requires both $v$ and $e$ to have equal types.

Rule \textsc{TRec} states that a recursive stream $\xrec{x}{e}$ has the recursive stream bound inside $e$.
The recursion must also be guarded, in that any recursive references to $x$ are delayed, but this requirement is performed as a separate syntactic check described in \autoref{s:core:causality}.

Rule \textsc{TCheck} states that checked properties $\xcheckP{\PStatus}{e}$ require a boolean property $e$.

Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ with a body expression of type $\tau$.
The overall expression has result type $\tau$.
Both rely and guarantee must be boolean expressions, and the guarantee can refer to the result as $x$.

\subsection{Dynamic semantics}
\label{s:core:dynamic}
\input{figures/core-bigstep.tex}

The dynamic semantics of Pipit are defined in \autoref{f:core-bigstep}.
We present our semantics in a big-step form.
This differs somewhat from traditional \emph{reactive} semantics of Lustre~\cite{caspi1995functional}.
Our big-step semantics emphasises the equational nature of Pipit, as it is substitution-based and syntax-directed, while the reactive semantics emphasises the finite-state streaming execution of the system.
We use transition systems for reasoning about the finite-state execution (\autoref{s:transition}), which is fairly standard~\cite{brun2023equation,champion2016kind2,raymond2008synchronous}.
Previous work on the {\sc W-calculus}~\cite{gallego2021w} for linear digital-signal-processing filters makes a similar distinction and provides a non-streaming semantics for reasoning about programs and a streaming semantics for executing programs.


The judgment form $\bigstep{\Sigma}{e}{v}$ denotes that expression $e$ evaluates to value $v$ under streaming history $\Sigma$.
The streaming history is a stream of heaps; in practice, we only evaluate expressions with a non-empty streaming history.

At a high level, evaluation unfolds recursive streams to determine a value.
For example, to evaluate the earlier sum example with input $\textit{ints} = [1; 2]$, we start with the judgment:
$$
\bigstep{\sumionetwo}{\sumrec}{v}
$$

First, we unfold the recursive stream one step to get $(\xfby{0}{\sumrec}) + \textit{ints}$.
Evaluation of primitives is standard.
To evaluate variables, we look for the variable in the current heap:
$$
\ruleIN{
}{
  \bigstep{\sumionetwo}{\textit{ints}}{2}
}{Var}
$$

For delays, we discard the current heap and continue evaluation with the history prefix:
$$
\ruleIN{
  \bigstep{\sgl{\textit{ints} \mapsto 1}}{\sumrec}{1}
}{
  \bigstep{\sumionetwo}{\xfby{0}{\sumrec}}{1}
}{$\mbox{Fby}_S$}
$$

Returning to \autoref{f:core-bigstep},
rule {\sc Var} evalutes a variable $x$ under some non-empty stream history $\Sigma; \sigma$, where $\sigma$ is the most recent heap.
Rules {\sc Value} and {\sc Let} are standard.
Rule {\sc Prim} evaluates a primitive $p$ applied to many arguments $e_1$ to $e_n$ by evaluating each argument separately; we then apply the primitive with prim-sem metafunction.
% rule {\sc Value} states that evaluating a value results in the value itself.
% Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then apply the primitive with prim-sem metafunction.
% Rule {\sc Let} is standard.

For delay expressions $\xfby{v}{e}$, we have two cases depending on whether there is a previous value.
When there is no previous value -- the streaming history only contains the current heap -- rule $\textsc{Fby}_1$ evaluates to the default value $v$.
Otherwise, rule $\textsc{Fby}_S$ applies; we evaluate the previous value of $e$ by discarding the most recent entry from the streaming history.

Rule {\sc Rec} evaluates a recursive stream $\xrec{x}{e}$ by unfolding the recursion one step.
For causal expressions (\autoref{s:core:causality}), where each recursive occurrence of $x$ is guarded by a followed-by, this unfolding eventually terminates as each followed-by shortens the history.

Rule {\sc Check} ignores the property when evaluating check expressions.
We do not dynamically check the property here; this is done in the checked semantics (\autoref{s:core:checked}).

Similarly, rule {\sc Contract} ignores preconditions and postconditions when evaluating contracts.
From an abstraction perspective, it would be valid to return an arbitrary value that satisfies the contract.
However, such an abstraction would make evaluation non-deterministic and, for contracts with unsatisfiable postconditions, non-total.
The deterministic and total nature of evaluation is key to our proofs and metatheory.

We also define two auxiliary judgment forms: $\bigsteps{\Sigma}{e}{V}$ and $\bigstepalways{\Sigma}{e}$.

Judgment form $\bigsteps{\Sigma}{e}{V}$ denotes that, under history $\Sigma$, expression $e$ evaluates to the \emph{stream} $V$.
This judgment performs iterated application of single-value evaluation.

Judgment form $\bigstepalways{\Sigma}{e}$ denotes that a boolean expression $e$ evaluates to the stream of trues under history $\Sigma$.
Informally, it can be read as ``$e$ is always true in history $\Sigma$''.

\subsection{Checked semantics}
\label{s:core:checked}

In addition to the big-step semantics above, we also define a judgment form for checking that the properties and contracts of a program hold for a particular streaming history.
We call these the \emph{checked} semantics; they are comparable to checking runtime assertions.
% Unlike an axiomatic semantics, the checked semantics operate on a concrete set of input streams.

The checked semantics have the judgment form $\semcheck{\Sigma}{\PStatus}{e}$, which denotes that under streaming history $\Sigma$, the properties and contracts of $e$ with status $\PStatus$ hold.
The property status dictates which properties should be checked and which should be ignored.

We consider a program to be \emph{valid} if its checks hold for all histories ($\forall \Sigma.~ \semcheck{\Sigma}{\PSValid}{e}$).
The checked semantics are a specification describing what it means to be a valid program.
We do not generally verify programs directly using the checked semantics; instead, we translate to an abstract transition system and construct the proofs there (\autoref{s:transition}).

% To show that an expression $e$'s unknown properties hold, we show that for all streaming histories $\Sigma$, the unknown properties hold ($\semcheck{\Sigma}{\PSUnknown}{e}$).
% When performing this proof, we can assume that the valid properties already hold to avoid re-checking properties ($\semcheck{\Sigma}{\PSValid}{e}$).
To check a property $(\xcheckP{\PStatus}{e})$ in history $\Sigma$, we check that $e$ is always true $(\bigstepalways{\Sigma}{e})$.

Checking contracts is more involved.
Conceptually, contracts involve two kinds of checks: one for the definition and one for the call-site.
To check that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid for some streaming history $\Sigma$, we check if, assuming the rely holds for the history ($\bigstepalways{\Sigma}{\erely}$), and assuming the body evaluates to some stream $V$ ($\bigsteps{\Sigma}{\ebody}{V}$), then the guarantee also holds ($\bigstepalways{\Sigma[x \mapsto V]}{\eguar}$).
% The fact that the checked semantics refers to a particular $\Sigma$ is significant: it allows the proof of contract validity to only consider streaming histories where the rely actually holds.

% and the unknown properties in the body and guarantee both hold ($\semcheck{\Sigma}{\PSUnknown}{\ebody} \wedge \semcheck{\Sigma}{\PSUnknown}{\eguar[x := \ebody]}$).
% and assuming that the valid properties in the rely, body and guarantee hold ($\semcheck{\Sigma}{\PSValid}{\erely} \wedge \semcheck{\Sigma}{\PSValid}{\ebody} \wedge \semcheck{\Sigma}{\PSValid}{\eguar[x := \ebody]}$)

For example, recall our earlier contract that the sum of strictly positive integers is positive:
$$
@let@~\text{sum}~\textit{i} = \sumcontractX{\PSUnknown}{i}
$$
To check the contract definition on a concrete input $\textit{i} = [1; 2]$, we first evaluate the body:
$$
\bigsteps{\sumionetwo}{\sumrecX{i}}{[1; 3]}
$$
We then check that, assuming all inputs are positive, then all results are positive:
$$
\bigstepalways{\sumionetwo}{\textit{i} > 0} \implies 
\bigstepalways{\sgl{\textit{i} \mapsto 1, \sumvar \mapsto 1}; \sgl{\textit{i} \mapsto 2, \sumvar \mapsto 3}}{sum > 0}
$$

It is critical that the rely is true \emph{at all points} in the stream.
Consider if we had instead used the input stream $\textit{i} = [-10; 1]$; here, the rely is false at the first step, but is instantaneously true at the second step.
In this case, the sum is $-10$ at the first step, and $-9$ at the second step.
At both steps the output is negative and the guarantee is false, even though the rely becomes true at the second step.
The contract itself remains valid, as the input did not satisfy the rely at all steps.
This behaviour is a key difference between streaming contracts and imperative pre-post contracts, which consider their preconditions in isolation.

Checking a contract instance is simple, in comparison: we just check that the call-site satisfies the rely at all steps.

% Contract instantiations often involve feedback loops, where the input to a contract depends on previous output values of the contract itself.
% For example, we can compute the Fibonacci sequence by instantiating the sum contract with its output fed back to it\footnote{The program shown here does not strictly fit within the core grammar, which does not support named functions; in the implementation we use the \fstar{} meta-language's top-level definitions for such definitions.}:
% $$
%   % @rec@~\fibvar.\, \sumcontractX{\PSUnknown}{(\xfby{1}{\fibvar})}
%   @rec@~\fibvar.\, \text{sum}~(\xfby{1}{\fibvar})
% $$

% To prove this instantiation valid, we need to show that the input $(\xfby{1}{\fibvar})$ is positive.
% To do this, we perform induction on the input streams, allowing us to assume that the rely was previously true.
% We then use the contract to show that, if the rely was true at the previous step, then the guarantee was true at the previous step -- which implies that the previous result is positive.

% To prove that a contract \emph{instantiation} (a call-site) is valid, we show that, under the calling environment, the rely clause is always true.
% Crucially, the proof can also use the fact that, if the rely is always true, then the guarantee is always true.
% This sort of feedback is necessary for proving properties of mutually-dependent calls.
% This circular dependency is well-founded as our causality check ensures that recursive streams are guarded by delays (\autoref{s:core:causality}).
% For example, the non-causal program $@rec@~\textit{nope}.\, \text{sum}~(-\textit{nope})$ depends instantaneously on itself.
% The rely requires that the negated result is positive; the guarantee ensures that the result is positive.
% However, there is no positive number whose negation is also positive, which suggests that the contract itself is inconsistent.
% This potential issue is resolved by outlawing such non-causal programs.

\input{figures/core-check.tex}

The checked semantics of Pipit is defined in \autoref{f:core-check}.
% The checked semantics descends into the expression and checks the properties and contracts as they are encountered.
% For binding forms such as let-expressions and recursive-expressions, the semantics evaluates the corresponding expression and extends the streaming history with the bound variable.

Rules {\sc ChkValue} and {\sc ChkVar} state that values and variables are always valid.

Rule {\sc ChkPrim} checks a primitive application by descending into the subexpressions.
Similarly, rule {\sc ChkFby} descends into followed-by expressions.

Rule {\sc ChkRec} checks a recursive-expression $\xrec{x}{e}$ by evaluating the overall expression to a stream of values $V$.
The rule then extends the streaming environment $\Sigma$ with $x$ bound to the values from $V$; this extended environment is used to descend into the recursive expression.

Rule {\sc ChkLet} checks a let-expression $\xlet{x}{e}{e'}$ descends into both sub-expressions.
To check the body $e'$, the rule first evaluates $e$ and extends the streaming environment.

Finally, the heavy lifting is performed by rules {\sc ChkCheck} and {\sc ChkContract}.

Rule {\sc ChkCheck} checks the properties marked $\PStatus$ in an expression $\xcheckP{\PStatus'}{e}$.
If the check-expression has the same status as what we are checking ($\PStatus = \PStatus'$), then we evaluate the expression $e$ and require it to be true at all steps.
We then unconditionally descend into the subexpression to check any nested properties.
Such nested properties are unlikely to be written directly by the user, but might occur after inlining.

Rule {\sc ChkContract} applies when checking property status $\PStatus$ of a contract with expression $\xcontractP{\PStatus'}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$.
This rule checks both the contract definition and the call-site.
We evaluate the body to a stream $V$; these values are used to check that the body satisfies guarantee.
Although the contract only has one property status, conceptually there are two distinct properties: one for the caller ($\PStatus'$) and one for the definition (assumed to be $\PSValid$).
To check the caller property when $\PStatus = \PStatus'$, we evaluate the rely $\erely$ and require it to hold.
To check the definition property when $\PStatus = \PSValid$, we assume that the rely holds, and check that the body satisfies the guarantee.
We also descend into the subexpressions to check them; when checking the body and guarantee, we can assume that the rely holds.
% This rule must deal with the two different roles of a contract at once; in the next section, we will separate the two roles.

% The semantics for contracts is fairly subtle, but we believe it is simpler than the translation to transition systems.
% Having a relatively simple high-level semantics of checks and contracts leaves room for future optimisations in translation to abstract transition systems.
% Additionally, existing contract support in Kind2~\cite{champion2016kind2} has rather ad-hoc restrictions, such as disallowing calling other contracts inside contract relies and guarantees. Our semantics handles this case without any special treatment.

\subsubsection{Blessing expressions and contracts}
\label{s:core:blessing}

Blessing is a meta-operation that replaces the property statuses in an expression so that all checks and contracts are marked as valid ($\PSValid$).
% Blessing an expression requires a proof that the unknown checks hold:
Blessing an expression requires a proof that, for all input streams, assuming the valid checks hold, then the unknown checks hold:

$$
\ruleIN{
  \forall \Sigma.~
  \semcheck{\Sigma}{\PSValid}{e}
  \implies
  \semcheck{\Sigma}{\PSUnknown}{e}
}{\text{bless}~e}{BlessExpression}
$$

We generally prove the required properties via abstract transition systems (\autoref{s:transition}).

Blessing is different for contract definitions, as we need to separate the definition of the contract from the instantiation.
To check that a contract definition is valid, we show that if the rely clause is always true for a particular input, then the body satisfies the guarantee for the same inputs.
We also assume that the valid properties in the rely, body and guarantee hold, and show the corresponding unknown properties:

\begin{tabbing}
  @MM@\= @MMMM@ \= @MMMMMMMMMMMMM@ \= \kill
  @let@ $\text{contract_valid}~\{ \erely \} ~\ebody~ \{ \eguar \}: \text{prop}$ = \\
  \> $\forall \Sigma.$
  \> $ (
    \semcheck{\Sigma}{\PSValid}{(\erely, \ebody, \eguar[x := \ebody])}
    ~\wedge~
    \bigstepalways{\Sigma}{\erely}
  ) $ \\
  \> $\implies$
  \> $(
    \semcheck{\Sigma}{\PSUnknown}{(\erely, \ebody, \eguar[x := \ebody])}
    ~\wedge~
    \bigstepalways{\Sigma}{\eguar[x := \ebody]}
    )$
\end{tabbing}

% XXX: the above uses substitution, but should use this bigsteps version. however, the above is much neater while below splits over the page.
% they are equivalent here anyway, so why not use the neater one?
% \begin{tabbing}
%   @MM@\= @MMMMM@ \= @MMMMMMMMMMMMM@ \= \kill
%   @let@ $\text{contract_valid}~\{ \erely \} ~\ebody~ \{ \eguar \}: \text{prop}$ = \\
%   \> $\forall \Sigma~V.$
%   \> $ (
%     \semcheck{\Sigma}{\PSValid}{(\erely, \ebody)}
%     ~\wedge~
%     \semcheck{\Sigma[x := V]}{\PSValid}{\eguar}
%     ~\wedge~
%     \bigstepalways{\Sigma}{\erely}
%     ~\wedge~
%     \bigsteps{\Sigma}{\ebody}{V}
%   ) $ \\
%   \> $\implies$
%   \> $(
%     \semcheck{\Sigma}{\PSUnknown}{(\erely, \ebody)}
%     ~\wedge~
%     \semcheck{\Sigma[x := V]}{\PSUnknown}{\eguar}
%     ~\wedge~
%     \bigstepalways{\Sigma}{\eguar}
%     )$
% \end{tabbing}



After proving that the contract is valid for all inputs, we can bless the contract definition.
Blessing the contract definition blesses the subexpressions for the rely, body and guarantee, but leaves the contract's \emph{instantiation} property status as unknown:
$$
\inferrule{
  \text{contract_valid}~\{ \erely \} ~\ebody~ \{ \eguar \}
}{\text{bless_contract}~\{\erely\}~\ebody~\{ \eguar\}}(\textsc{BlessContract})
$$

% The contract we saw in \autoref{s:motivation:contract} did not explicitly bless its contract, but the 


\subsection{Causality and metatheory}
\label{s:core:causality}

To ensure that recursive streams have a computational interpretation, we implement a causality restriction, similar to standard Lustre~\cite{caspi1995functional}.
This restriction checks that all recursive streams are guarded by a followed-by delay.
We implement this as a simple syntactic check: each $\xrec{x}{e}$ can only mention $x$ inside a followed-by.
This check is stricter than necessary: for example, the expression $\xrec{x}{(\xlet{x'}{x + 1}{\xfby{0}{x'}})}$ does mention the recursive stream $x$ outside of the delay, but after inlining the let, it would be causal.
We hope to relax this restriction somewhat in future work.

The causality restriction gives us some important properties about the metatheory.
The most important property is that the dynamic semantics form a total function: given a streaming history and a causal expression, we can evaluate the expression to a value.
These properties are mechanised in \fstar{}.

% \begin{theorem}[bigstep-deterministic]
%   For any streaming history $\Sigma$, if expression $e$ evaluates to $v$ $(\bigstep{\Sigma}{e}{v})$ and also evaluates to $v'$ $(\bigstep{\Sigma}{e}{v'})$, then $v = v'$.
% \end{theorem}

\begin{theorem}[bigstep-is-total]
  For any non-empty streaming history $\Sigma$ and causal expression $e$, there exists some value $v$ such that $e$ evaluates to $v$ $(\bigstep{\Sigma}{e}{v})$.
\end{theorem}

The relationship between substitution and the streaming history is also important.
In general, we have a substitution property that states that evaluating a substituted expression $e[x := e']$ under some context $\Sigma$ is equivalent to evaluating $e'$ and adding it to the context $\Sigma$:

\begin{theorem}[bigstep-substitute]
  For a streaming history $\Sigma$ and causal expressions $e$ and $e'$, if $e[x := e']$ evaluates to a value $v$ $(\bigstep{\Sigma}{e}{v})$, then we can evaluate $e'$ to some stream $V$ $(\bigsteps{\Sigma}{e'}{V})$ and extend the streaming history to evaluate $e$ to the original value $(\bigstep{\Sigma[x \mapsto V]}{e}{v})$.
  The converse is also true.
\end{theorem}

% \begin{theorem}[bigstep-rec-substitute-elim]
%   For a streaming history $\Sigma; \sigma$ and a causal recursive expression $\xrec{x}{e}$, if the recursive expression evaluates to a stream $(\bigsteps{\Sigma; \sigma}{\xrec{x}{e}}{V; v})$, then we can evaluate $e$ to the same value by binding $x$ to the same stream in the streaming history $(\bigstep{(\Sigma; \sigma)[x \mapsto (V; v)]}{e}{v})$.
% \end{theorem}

The big-step semantics in \autoref{f:core-bigstep} for a recursive expression $\xrec{x}{e}$ performs one step of recursion by substituting $x$ for the recursive expression.
An alternative non-syntax-directed semantics would be to have the environment outside the semantics supply a stream $V$ such that if we extend the streaming history with $x \mapsto V$, then $e$ evaluates to $V$ itself.
The above substitution theorem can be used to show that, for causal expressions, these two semantics are equivalent.
We can additionally show that, when evaluating $e$ with $x \mapsto V$, the most recent value in $V$ does not affect the result.
This fact can be used to ``seed'' evaluation by starting with an arbitrary value:
\begin{theorem}[bigstep-rec-causal]
  For a streaming history $\Sigma; \sigma$ and a causal recursive expression $\xrec{x}{e}$, if $(\bigstep{\Sigma; \sigma}{e}{v})$, then updating $\sigma[x]$ with any value $v'$ results in the same value: $(\bigstep{\Sigma; \sigma[x \mapsto v']}{e}{v})$.
\end{theorem}

