%!TEX root = ../Main.tex

\section{Conclusion}

We have presented Pipit, a verified compiler and proof system for reactive systems.
Our implementation of the TTCAN driver logic shows that Pipit can express programs which are currently unsupported by other verified Lustre compilers.
Pipit can also verify high-level program properties which are difficult to express and prove in existing Lustre model-checkers.
Our development includes verified translations to both abstract and executable transition systems; both are shown to preserve the dynamic semantics.
We also introduced a checked semantics, which describes the proof obligations of a program; in future work, we intend to verify that any translated property the user proves on the transition system holds for the original property on the checked semantics.

In the future, we intend to increase the expressivity of Pipit by adding \emph{clocks}, which are used to describe partially-defined streams~\cite{caspi1995functional}.
Clocks are important for composing complex systems together and avoiding unnecessary computation.

We are interested in further pursuing the intersection of model-checking with interactive theorem proving.
A smart-contract called Djed \cite{zahnentferner2023djed} currently uses a mixture of Kind2 \cite{champion2016kind2} and manual Isabelle/HOL proofs to show that the contract is well-behaved.
In future work, we would like to further investigate whether Pipit's integration of streaming proofs with \fstar{}'s automated proof system would be able to provide similar proofs, without introducing any semantic gap between the two systems.
